\texttt{apt-p2p}: A Peer-to-Peer Distribution System for Software Package Releases and Updates

### 2008-09-28 docs/paper/paper.tex
Final version of paper.

In this paper, we propose a novel peer-to-peer assisted distribution system design that
addresses the above challenges. It enhances the existing distribution systems by providing compatible and yet more efficient downloading and updating services
for software packages. Our design leads to \texttt{apt-p2p}, a practical implementation that extends the popular \texttt{apt} distributor.  \texttt{apt-p2p}  has been used in conjuction with Debian-based distribution of Linux
software packages and will also be available in the next release of Ubuntu. We have addressed the key design issues in \texttt{apt-p2p}, including indexing table customization,
response time reduction, and multi-value extension. They together ensure
that the altruistic users' resources are effectively utilized and thus significantly reduces the currently
large bandwidth requirements of hosting the software, as confirmed by our existing real user statistics.
\end{abstract}

@@

With the widespread penetration of broadband access, the Internet has become a cost-effective
vehicle for software development and release. This is particularly true for the free software
community whose developers and users are distributed worldwide and work asynchronously. The ever increasing power of
modern programming languages, computer platforms, and operating systems has made this software extremely large and complex,
though it is often divided into a huge number of small packages.
Together with their popularity among users,
an efficient and reliable management and distribution of these software packages over the Internet has become a challenging task.

The existing distribution for free software is mostly based on the client/server model, e.g.,
the Advanced Package Tool (\texttt{apt}) for Linux, which suffers from the well-known bottleneck problem.
Given the free nature of this software, there are often a number of users
motivated by altruism to help out with the distribution, so as to promote the healthy development
of this voluntary society.
We thus naturally expect that peer-to-peer distribution can be implemented in
this context, which will scale well with the currently large user bases and can easily explore the network resources made available by
the volunteers.

@@

In the free software society, there are a large number of groups using the Internet to
collaboratively develop and release their software. The ever increasing power of
modern programming languages and operating systems has made this software, like commercial software, extremely large and complex, though it is often
distributed in many small units (packages). Together with their popularity among users,
an efficient and reliable management and distribution of these packages over the Internet has become a daunting task. In this section, we offer concrete examples illustrating the
unique challenges in this context.

@@

The software packages being distributed are being constantly
updated. These updates could be the result of the software creators
releasing a new version with improved functionality,
or the distributor updating their packaging of the
software to meet new requirements. Even if the distributor
periodically makes \emph{stable} releases, which are snapshots of
all the packages in the archive at a certain time, later updates are still
released for security issues or serious bugs.

@@

For example, Figure~\ref{update_size} shows the amount of data in
the Debian archive that was updated each day over a period of 3
months. Every single day, approximately 1.5\% of the 119,000 MB archive is
updated with new versions of packages. This frequency is much higher than
that of most commercial software, mainly because much of free software is
developed in a loosely managed environment of developers working
asynchronously on a worldwide scale.

@@

Finally, though there are a large number of packages and a large number of
users, the interest in a particular package, or version of a package, can be very
limited. Specifically, there are core packages that every user has to download, but most
packages would fall in the category of optional or extra, and so are
interesting to only a limited number of people.

@@

Given the relatively long time for software package downloading, existing package management systems generally display
some kind of indication of speed and completeness for users to
monitor. Since previous client-server downloads occurred in a sequential
fashion, the package management software also measures the speed
based on sequential downloading. To offer comparable user experience, it is natural to expect that
the new peer-to-peer solution be reasonably responsive at retrieving packages,
preferably in a sequential downloading order too.

\subsection{Why BitTorrent Doesn't Work Well}
\label{bittorrent}

@@

Our model for using peer-to-peer to enhance package distribution is shown in
Figure~\ref{model}. As shown in Phase~1, our program will act as a
proxy (1,2), downloading (3) and caching all files communicated
between the user and the server (4). It will therefore also have
available the index files containing the cryptographic hashes all
packages. Later, in Phase~2, upon receiving a request from the user
to download a package (5), our program will search the index files
for the package being requested and find its hash (6). This hash can
then be looked up recursively in an indexing structure (a Distributed Hash Table in our implementation) (7), which will return a
list of peers that have the package already (8). The package can
then be downloaded from the peers (11,12), it can be verified using
the hash (13), and if valid can be returned to the user (14). The
current nodes location is also added to the DHT for that hash (15),
as it is now a source for others to download from.

In steps (11,12), the fact that this package is also available to download for free
from a server is very important to our proposed model. If the package hash
can not be found in the DHT, the peer can then fallback to
downloading from the original location (i.e. the server).
The server thus, with no modification to its
functionality, serves as a seed for the packages in the peer-to-peer
system. Any packages that have just been updated, or that are very
rare, and so don't have any peers available, can always be found on
the server. Once the peer has completed the download from the server
and verified the package, it can then add itself to the DHT as the
first peer for the new package, so that future requests for the package
will not need the server.

This sparse
interest in a large number of packages undergoing constant updating
is well suited to the functionality provided by a Distributed Hash
Table (DHT). DHTs require unique keys to store and retrieve strings
of data, for which the cryptographic hashes used by these package
management systems are perfect for. The stored and retrieved strings
can then be pointers to the peers that have the package that hashes
to that key.

@@

% \subsection{Implementation Options}
% \label{imp_options}
% 
% There are several ways to implement the desired peer-to-peer functionality
% into the existing package management software. The functionality can
% be directly integrated through modifications to the software, though
% this could be difficult as the peer-to-peer functionality should be running
% at all times. This is needed both for efficient lookups and to
% support uploading of already downloaded packages. Unfortunately, the
% package management tools typically only run until the download and
% install request is complete.
% 
% Many of the package management software implementations use HTTP
% requests from web servers to download the packages, which makes it
% possible to implement the peer-to-peer aspect as an almost standard HTTP
% caching proxy. This proxy will run as a daemon in the background,
% listening for requests from the package management tool for package
% files, as well as serving (uploading) cached package to other peers.
% It will get uncached requests first from the peer-to-peer system, or
% fall back to the normal HTTP request from a server should it not
% be found. For methods that don't use HTTP requests, other types of
% proxies may also be possible.

@@

Our \texttt{apt-p2p} implementation supporting the Debian package
distribution system has been available to all Debian users since May
3rd, 2008 \cite{apt-p2p-debian}, and will also be available in the
next release of Ubuntu \cite{apt-p2p-ubuntu}. We have since created
a \emph{walker} that will navigate the DHT and find all the peers
currently connected to it. This allows us to analyze many aspects of
our implementation.

@@

There are a number of works dedicated to developing a collaborative
content distribution network (CDN) using peer-to-peer techniques.
Freedman et. al. developed Coral \cite{coral} using a distributed
\emph{sloppy} hash table to speed request times. Pierre and van
Steen developed Globule \cite{globule} which uses typical DNS and
HTTP redirection techniques to serve requests from a network of
replica servers, which in turn draw their content from the original
location (or a backup). Shah et. al. \cite{shah08} analyze an
existing software delivery system and use the results to design a
peer-to-peer content distribution network that makes use of
volunteer servers to help with the load. None of these systems meets
our goal of an even distribution of load amongst the users of the
system. Not all users of the systems become peers, and so are not
able to contribute back to the system after downloading. The
volunteers that do contribute as servers are required to contribute
larger amounts of bandwidth, both for uploading to others, and in
downloading content they are not in need of in order to share them
with other users. Our system treats all users equally, requiring all
to become peers in the system, sharing the uploading load equally
amongst all, but doesn't require any user to download files they
would not otherwise need.

The most similar works to ours are by Shah et. al. \cite{shah08} and
Shark by Annapureddy et. al. \cite{shark}.
Shah's system, in addition to the drawbacks mentioned above,
is not focused on the interactivity of downloads, as
half of all requests were required ``to wait between 8 and 15
minutes.'' In contrast, lookups in our system take only seconds to
complete, and all requests can be completed in under a minute.
Shark makes use of Coral's distributed sloppy hash table to speed
the lookup time, but their system is more suited to its intended use
as a distributed file server. It doesn't make use of authoritative
copies of the original files, allowing instead any users in the
system to update files and propagate those changes to others. Our
system is well-tailored to the application of disseminating the
unchanging software packages from the authoritative sources to all
users.

>>>>

In this paper, we propose a novel peer-to-peer assisted distribution system design that
addresses the above challenges. It enhances the existing distribution systems by providing compatible and yet more efficient downloading and updating services
for software packages. Our design leads to \texttt{apt-p2p}, a practical implementation that extends the popular \texttt{apt} distributor.  \texttt{apt-p2p}  has been used in conjuction with Debian-based distribution of Linux
software packages and will also be available in the next release of Ubuntu. We have addressed the key design issues in \texttt{apt-p2p}, including indexing table customization,
response time reduction, and multi-value extension. They together ensure
that the altruistic users' resources are effectively utilized and thus significantly reduces the currently
large bandwidth requirements of hosting the software, as confirmed by our existing real user statistics over the Internet.
\end{abstract}

@@

With the widespread penetration of broadband access, the Internet has become a cost-effective
vehicle for software development and release \cite{feller2000fao}. This is particularly true for the free software
community whose developers and users are distributed worldwide and work asynchronously. The ever increasing power of
modern programming languages, computer platforms, and operating systems has made this software extremely large and complex,
though it is often divided into a huge number of small packages.
Together with their popularity among users,
an efficient and reliable management and distribution of these software packages over the Internet has become a challenging task \cite{ubuntu-blueprint}.

The existing distribution for free software is mostly based on the client/server model, e.g.,
the Advanced Package Tool (\texttt{apt}) for Linux \cite{apt}, which suffers from the well-known bottleneck problem.
Given the free nature of these software, there are often a number of users
motivated by altruism to help out with the distribution, so as to promote the healthy development
of this voluntary society.
We thus naturally expect that peer-to-peer distribution can be implemented in
this context, which will scale well with the currently large user bases and can easily explore the resources made available by
the volunteers.

@@

In the free software community, there are a large number of groups using the Internet to
collaboratively develop and release their software. Efficient and reliable management and distribution 
of these software packages over the Internet thus has become a critical task. In this section, we offer concrete examples illustrating the
unique challenges in this context.

@@

The software packages being distributed are being constantly
updated. These updates could be the result of the software creators
releasing a new version with improved functionality,
or the distributor updating their packaging of the
software to meet new requirements. Even if the distributor
periodically makes \emph{stable} releases, which are snapshots of
all the packages in the archive at a certain time, many later updates are still
released for security issues or serious bugs.

@@

For example, Figure~\ref{update_size} shows the amount of data in
the Debian archive that was updated each day over a period of 3
months. Every single day, approximately 1.5\% of the 119,000 MB archive is
updated with new versions of packages. This frequency is much higher than
that of most commercial software \cite{microsoft-update}, mainly because much of free software is
developed in a loosely managed environment of developers working
asynchronously on a worldwide scale.

@@

Though there are a large number of packages and a large number of
users, the interest in a particular package, or version of a package, can be very
limited. Specifically, there are core packages that every user has to download, but most
packages would fall in the category of optional or extra, and so are
interesting to only a limited number of users.

@@

Finally, given the relatively long time for software package downloading, existing package management systems generally display
some kind of indication of speed and completeness for users to
monitor. Since previous client-server downloads occurred in a sequential
fashion, the package management software also measures the speed
based on sequential downloading. To offer comparable user experience, it is natural to expect that
the new peer-to-peer solution be reasonably responsive at retrieving packages,
preferably in a sequential downloading order too.

\subsection{Why BitTorrent Does Not Work Well}
\label{bittorrent}

@@

\subsection{System Overview}
Our model for using peer-to-peer to enhance package distribution is shown in
Figure~\ref{model}. As shown in Phase~1, our program will act as a
proxy (1,2), downloading (3) and caching all files communicated
between the user and the server (4). It will therefore also have
available the index files containing the cryptographic hashes all
packages. Later, in Phase~2, upon receiving a request from the user
to download a package (5), our program will search the index files
for the package being requested and find its hash (6). This hash can
then be looked up recursively in an indexing structure (a Distributed Hash Table, or DHT \cite{kademlia}, in our implementation) (7), which will return a
list of peers that have the package already (8). The package can
then be downloaded from the peers (11,12), it can be verified using
the hash (13), and if valid can be returned to the user (14). The
current nodes location is also added to the DHT for that hash (15),
as it is now a source for others to download from.

In steps (11,12), the fact that this package is also available to download for free
from a server is very important to our proposed model. If the package hash
can not be found in the DHT, the peer can then fallback to
downloading from the original location (i.e. the server).
The server thus, with no modification to its
functionality, serves as a seed for the packages in the peer-to-peer
system. Any packages that have just been updated, or that are very
rare, and so do not have any peers available, can always be found on
the server. Once the peer has completed the download from the server
and verified the package, it can then add itself to the DHT as the
first peer for the new package, so that future requests for the package
will not need the server.

This sparse
interest in a large number of packages undergoing constant updating
is well suited to the functionality provided by a DHT. A DHT requires unique keys to store and retrieve strings
of data, for which the cryptographic hashes used by these package
management systems are perfect for. The stored and retrieved strings
can then be pointers to the peers that have the package that hashes
to that key.

@@

% \subsection{Implementation Options}
% \label{imp_options}
%
% There are several ways to implement the desired peer-to-peer functionality
% into the existing package management software. The functionality can
% be directly integrated through modifications to the software, though
% this could be difficult as the peer-to-peer functionality should be running
% at all times. This is needed both for efficient lookups and to
% support uploading of already downloaded packages. Unfortunately, the
% package management tools typically only run until the download and
% install request is complete.
%
% Many of the package management software implementations use HTTP
% requests from web servers to download the packages, which makes it
% possible to implement the peer-to-peer aspect as an almost standard HTTP
% caching proxy. This proxy will run as a daemon in the background,
% listening for requests from the package management tool for package
% files, as well as serving (uploading) cached package to other peers.
% It will get uncached requests first from the peer-to-peer system, or
% fall back to the normal HTTP request from a server should it not
% be found. For methods that do not use HTTP requests, other types of
% proxies may also be possible.

@@

Our \texttt{apt-p2p} implementation supporting the Debian package
distribution system has been available to all Debian users since May
3rd, 2008 \cite{apt-p2p-debian}, and will also be available in the
next release of Ubuntu \cite{apt-p2p-ubuntu}. We have since created
a \emph{walker} that will navigate the DHT and find all the peers
currently connected to it. This allows us to analyze many aspects of
our implementation in the real Internet environment.

@@

There are a number of works dedicated to developing a collaborative
content distribution network (CDN) using peer-to-peer techniques.
Freedman et. al. developed Coral \cite{coral} using a distributed
\emph{sloppy} hash table to speed request times. Pierre and van
Steen developed Globule \cite{globule} which uses typical DNS and
HTTP redirection techniques to serve requests from a network of
replica servers, which in turn draw their content from the original
location (or a backup). Shah et. al. \cite{shah08} analyze an
existing software delivery system and use the results to design a
peer-to-peer content distribution network that makes use of
volunteer servers to help with the load. None of these systems meets
our goal of an even distribution of load amongst the users of the
system. Not all users of the systems become peers, and so are not
able to contribute back to the system after downloading. The
volunteers that do contribute as servers are required to contribute
larger amounts of bandwidth, both for uploading to others, and in
downloading content they are not in need of in order to share them
with other users. Our system treats all users equally, requiring all
to become peers in the system, sharing the uploading load equally
amongst all, but does not require any user to download files they
would not otherwise need.

The most similar works to ours are by Shah et. al. \cite{shah08} and
Shark by Annapureddy et. al. \cite{shark}.
Shah's system, in addition to the drawbacks mentioned above,
is not focused on the interactivity of downloads, as
half of all requests were required ``to wait between 8 and 15
minutes.'' In contrast, lookups in our system take only seconds to
complete, and all requests can be completed in under a minute.
Shark makes use of Coral's distributed sloppy hash table to speed
the lookup time, but their system is more suited to its intended use
as a distributed file server. It does not make use of authoritative
copies of the original files, allowing instead any users in the
system to update files and propagate those changes to others. Our
system is well-tailored to the application of disseminating the
unchanging software packages from the authoritative sources to all
users.
